\Large{Adversarial Training against Systematic Uncertainty}

### 2016-09-27 tex/paper.tex
bibliography style
save

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Etiam commodo, enim
vitae facilisis pretium, ligula justo aliquet lectus, non interdum erat lorem et
nisi. Donec pharetra lectus in magna pellentesque vehicula. Praesent dapibus
lorem sed enim lacinia, a vestibulum sapien mattis. Sed ornare mollis aliquet.
Nulla tempor lacinia tortor, in rhoncus augue porta nec. Morbi sed convallis
nibh, eu hendrerit turpis. Curabitur sit amet rhoncus purus. Curabitur eget
magna lorem. Phasellus pretium nisi quis est tincidunt, faucibus vulputate augue
viverra. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Proin a urna a
ex egestas pulvinar.

@@

The discovery of new particles or physical phenomena in high energy physics
experiments, like those at the Large Hadron Collider, requires the observation
of statistically significant deviations from the predictions of the Standard
Model, the current model of the known fundamental particles and their
interactions.  Typically, discovery requires satisfying the \textit{5-sigma
rule}, whereby the deviation must be at least five standard deviations from
predictions, i.e. having a p-value of $p \lesssim 3 \times 10^{-7}$. The
evaluation of the statistical significance must not only account for statistical
uncertainties \glnote{Shouldnt we define what are statistical uncertainties, so
that systematic uncertainties are in constrast better understood?}, but also
systematic uncertainties that represent our incomplete knowledge of the theory
of particle interactions or of the properties of the experimental detection
apparatus. Systematic uncertainties can alter the expected rate of signal and
background classes, e.g. their priors, as well as the distributions of features.
As such, the signal class true-positive rate and the background class
false-positive rate of a classifier can change with variations in the systematic
uncertainty thereby decreasing the sensitivity of an analysis. In this paper, we
present a new strategy for training a classifier, based on adversarial
training~\citep{goodfellow2014generative}, which aims to build a classifier
whose decision function is invariant under variations due to systematic
uncertainties.  In this way, uncertainties on predictions from simulations will
have a reduced impact when computing the significance of potential deviations
after application of the discriminant model and thereby increase the sensitivity
of the analysis to signals from potential new physics sources. \glnote{Shall we
be more explicit somewhere and put a formula underlining the effect of having a
pivotal classifier on the resulting sensitivity? (versus a classifier whose
decision function depends on the nuisance.)}

Given data $X$ and associated labels $Y$
taking values $y \in {\cal Y} = \{s,
b_{i=1...k} \}$ where $s$ is the signal class and $b_{i=1...k}$ are
the background classes, then the probability density of the data
can be written as a mixture
\begin{equation}
p(X) = \pi_s p(X|Y=s) + \sum_{i=1}^{k} \pi_{b_i} p(X | Y=b_i),
\end{equation}
where $\pi_{y}$ represents the prior for the signal or a given
background, and $p(X|Y)$ is the conditional distribution of the data
for the signal or a given background. Systematic uncertainties can
affect this distribution in the following ways \glnote{add references}:
\begin{itemize}
\item Uncertainties on the priors $\pi_{y}$.  These arise from
  uncertainties on the rates we expect a given process to occur as
  predicted from theoretical calculations or from alternative
  measurements of the rate of a given class.

@@

Systematic uncertainties are parameterized by nuisance parameters in the
statistical analysis of data \glnote{add references}.  Nuisance parameter often,
though not always, have a suitable prior coming from alternative measurements in
data.  For instance, The average measured energy of an electron in simulation
may be compared with those in data using a relatively pure sample of $Z$ boson
decays to electrons.  The average measured electron energy can not be measured
without uncertainty, and the measured uncertainty can be used in other analyses
to estimate the uncertainty deriving from our imperfect knowledge of the
electron's energy.  A nuisance parameter is assigned to the electron energy
uncertainty, which controls the variation of the electron energy in the
statistical analysis of data.  Nuisance parameters are typically, though not
always, constrained by a suitable prior, e.g. a Gaussian distribution with a
standard deviation equal to the measured uncertainty. \glnote{... and
accordingly, we want to build a classifier independent on the nuisance
parameter, as a way to be independent of the corresponding systematics.}

@@

    \caption{Architecture for the adversarial training of a binary classifier
    $f$ against a continuous parameter $Z$.}
    \label{fig:architecture}
    \end{figure*}

@@

\subsection{Toy example}

@@

\glnote{Define the architecture of $r$ when $Z$ is categorical ($r$ is a
standard classifier) or continuous ($r$ is e.g a 2-output NN such that the
distribution of $Z|f(X)$ is modeled by a truncated Gaussian of known support). Cite Nix.}

@@

To account for systematic uncertainties, experimentalists in high energy physics
typically take as fixed a classifier $f$ built from training data for a nominal
value $z_0$ of the nuisance parameter, and then propagate uncertainty
\glnote{add ref} by estimating $p(f(x)|z)$ with a parameterized calibration
procedure. Clearly, this classifier is however not optimal for $z \neq z_0$. In
this setting, parameterized
classifiers~\citep{cranmer2015approximating,Baldi:2016fzo} directly take
(nuisance) parameters as additional input variables, hence ultimately providing
the most statistically powerful approach for incorporating the effect of
systematics on the underlying classification task.  As argued in
\citep{Neal:2007zz}, such classifiers can however not be used on real data since
the correct value $z$ of the nuisance often remains unknown. This is typically
not an issue in the context of parameter
inference~\citep{cranmer2015approximating}, where nuisance parameters are
marginalized out, but otherwise often limits the range of their applications. In
practice, parameterized classifiers  are also computationally expensive to build
and evaluate. In particular, calibrating their decision function, i.e.
approximating $p(f(x,z)|z)$ as a continuous function of $z$, remains an open
challenge. By contrast, constraining $f$ to be pivotal yields a classifier which
may not be optimal with respect to $Y|X,Z$, as discussed in
Section~\ref{sec:theory}, but that can otherwise be used in a wider range of
applications, since knowing the correct value $z$ of the nuisance is not
necessary. Similarly, calibration needs to be carried out only once, since  the
dependence on the nuisance is now built-in. \glnote{Shall we also discuss the
relation with \citep{Neal:2007zz} where point estimates of the nuisance are used
as inputs to $f$?}

@@

\bibliography{bibliography.bib}

>>>>

Nice and cozy abstract goes here...

@@

The discovery of new particles or physical phenomena in high energy physics
experiments, like those at the Large Hadron Collider, requires the observation
of statistically significant deviations from the predictions of the Standard
Model, the current model of the known fundamental particles and their
interactions.  Typically, discovery requires satisfying the \textit{5-sigma
rule}, whereby the deviation must be at least five standard deviations from
predictions, i.e. having a p-value of $p \lesssim 3 \times 10^{-7}$. The
evaluation of the statistical significance must not only account for statistical
uncertainties due to random fluctuations, but also for
systematic uncertainties that represent our incomplete knowledge of the theory
of particle interactions or of the properties of the experimental detection
apparatus. Systematic uncertainties can alter the expected rate of signal and
background classes, e.g. their priors, as well as the distributions of features.
As such, the signal class true-positive rate and the background class
false-positive rate of a classifier can change with variations in the systematic
uncertainty thereby decreasing the sensitivity of an analysis. In this paper, we
present a new strategy for training a classifier, based on adversarial
training~\citep{goodfellow2014generative}, which aims at building a classifier
whose decision function is invariant under variations due to systematic
uncertainties.  In this way, uncertainties on predictions from simulations will
have a reduced impact when computing the significance of potential deviations
after application of the discriminant model and thereby increase the sensitivity
of the analysis to signals from potential new physics sources. \glnote{Shall we
be more explicit somewhere and put a formula underlining the effect of having a
pivotal classifier on the resulting sensitivity? (versus a classifier whose
decision function depends on the nuisance.)}

Given data $X$ and associated labels $Y$
taking values $y \in {\cal Y} = \{s,
b_{i=1...k} \}$ where $s$ is the signal class and $b_{i=1...k}$ are
the background classes, the probability density of the data
can be written as a mixture
\begin{equation}
p(X) = \pi_s p(X|Y=s) + \sum_{i=1}^{k} \pi_{b_i} p(X | Y=b_i),
\end{equation}
where $\pi_{y}$ represents the prior for the signal or a given
background, and $p(X|Y)$ is the conditional distribution of the data
for the signal or a given background. Systematic uncertainties can
affect this distribution in the following ways \glnote{add references}:
\begin{itemize}
\item Uncertainties on the priors $\pi_{y}$.  These arise from
  uncertainties on the rates we expect a given process to occur as
  predicted from theoretical calculations or from alternative
  measurements of the rate of a given class.

@@

Systematic uncertainties are parameterized by nuisance parameters in the
statistical analysis of data \glnote{add references}.  Nuisance parameter often,
though not always, have a suitable prior coming from alternative measurements in
data.  For instance, the average measured energy of an electron in simulation
may be compared with those in data using a relatively pure sample of $Z$ boson
decays to electrons.  The average measured electron energy can not be measured
without uncertainty, and the measured uncertainty can be used in other analyses
to estimate the uncertainty deriving from our imperfect knowledge of the
electron's energy.  A nuisance parameter is assigned to the electron energy
uncertainty, which controls the variation of the electron energy in the
statistical analysis of data.  Nuisance parameters are typically, though not
always, constrained by a suitable prior, e.g. a Gaussian distribution with a
standard deviation equal to the measured uncertainty. \glnote{... and
accordingly, we want to build a classifier independent on the nuisance
parameter, as a way to be independent of the corresponding systematics.}

@@

    \caption{Architecture for the adversarial training of a binary classifier
    $f$ against a nuisance parameter $Z$. The adversary $r$ models the
    distribution $p(Z|f(X;\theta_f))$ of the nuisance as observed only through the output $f(X;\theta_f)$ of the classifier. By
    maximizing the antagonistic objective ${\cal L}_r(\theta_f, \theta_r)$ (as part of minimizing ${\cal L}_f(\theta_f) - \lambda {\cal L}_r(\theta_f, \theta_r)$), the classifier
    $f$ forces $p(Z|f(X;\theta_f)$ to become uniform, which happens
    when $f(X;\theta_f)$ is  independent of the nuisance parameter $Z$ and therefore pivotal.}

    \label{fig:architecture}
\end{figure*}

@@

\subsection{Toy example}
\label{sec:toy}

@@

@@

To account for systematic uncertainties, experimentalists in high energy physics
typically take as fixed a classifier $f$ built from training data for a nominal
value $z_0$ of the nuisance parameter, and then propagate uncertainty
\glnote{add ref} by estimating $p(f(x)|z)$ with a parameterized calibration
procedure. Clearly, this classifier is however not optimal for $z \neq z_0$.
To overcome this issue, the classifier $f$ is sometimes \glnote{add refs} built instead on a mixture
of training data generated from several nominal values $z_0, z_1, \dots$ of the nuisance.
While this certainly improves with respect to classification performance,
there is however no guarantee that the resulting classifier is pivotal, as shown
previously in Section~\ref{sec:toy}.
As an alternative, parameterized
classifiers~\citep{cranmer2015approximating,Baldi:2016fzo} directly take
(nuisance) parameters as additional input variables, hence ultimately providing
the most statistically powerful approach for incorporating the effect of
systematics on the underlying classification task.  As argued in
\citep{Neal:2007zz}, such classifiers can however not be used on real data since
the correct value $z$ of the nuisance often remains unknown. This is typically
not an issue in the context of parameter
inference~\citep{cranmer2015approximating}, where nuisance parameters are
marginalized out, but otherwise often limits the range of their applications. In
practice, parameterized classifiers  are also computationally expensive to build
and evaluate. In particular, calibrating their decision function, i.e.
approximating $p(f(x,z)|z)$ as a continuous function of $z$, remains an open
challenge. By contrast, constraining $f$ to be pivotal yields a classifier which
may not be optimal with respect to $Y|X,Z$, as discussed in
Section~\ref{sec:theory}, but that can otherwise be used in a wider range of
applications, since knowing the correct value $z$ of the nuisance is not
necessary. Similarly, calibration needs to be carried out only once, since  the
dependence on the nuisance is now built-in.

@@

\bibliographystyle{acm}
\bibliography{bibliography.bib}
